\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-11 analysis/paper/paper.Rmd

---
documentclass: jss
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 7
fig_height: 6
fig_caption: true
---

@@

library(tidyverse)
library(magrittr)
library(survival)
library(casebase)
library(splines)
#library(readr)
#library(tibble)
library(glmnet)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(2)

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- FALSE
eval_cs3 <- FALSE
eval_cs4 <- TRUE
```

@@

At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [-@survTaskView]. For the purposes of this article, we restricted our description to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and cumulative incidence curve estimation. By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were further reduced through manual curation to a subset of 14 packages. These packages appears in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer. \textcolor{red}{JESSE: It's not clear how you selected this subset, going from 60 to 14 is a big drop.}

Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche]. Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: \pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package]. On the other hand, \pkg{casebase} provides penalized estimation in the context of parametric hazards. To our knowledge, this is the only package to offer this functionality. \textcolor{red}{JESSE: Is this true? Because it seems to be a logical conclusion from your paragraph.}

@@

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks. The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression. \textcolor{red}{JESSE: There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}. Could you incorporate some of their discussion, with proper citations? The biggest omission right now is \pkg{cmprsk}, which we actually use below.}

@@

```{r echo = FALSE}
CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group",])
CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group",])
CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group",])

@@

We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [-@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital. Imputation was conducted using the default methods in the \pkg{mice} package in \proglang{R}, available on CRAN. \textcolor{red}{Why was this method used? The Vanderbilt University website lists describes another method for imputation that we could have used.} Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest. The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual. A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}. The data was downloaded from the main author's website, and it is also available as part of the \pkg{casebase} package.

@@

Hospital death as a covariate was removed, as this is directly informative of death. The \pkg{glmnet} package on CRAN was used to introduce elastic net, lasso and ridge regressions to \pkg{casebase}. To visualize the effect of imputation on incidence density, we can refer to the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and post-imputation population time plot in Figure \ref{fig:ptPostI}. The incidence density is consistent between imputed and complete datasets.

```{r supportSetUp, echo = FALSE, eval = eval_cs3}
completeData <- casebase::support
#keep one individual out, at random, for which we will develop an 
#absolute risk curve.
sam <- sample(1:nrow(completeData), 1)
#Set workingData to the remaining individuals in the data
#workingData=completeData[-c(sam),]
workingCompleteData <- as.matrix(sparse.model.matrix(death~ . - d.time, data = completeData))[,-c(1)]
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData <- workingCompleteData[sam,]
x <- workingCompleteData[-c(sam),]
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y <- data.matrix(completeData[-c(sam), c(34, 4)])
```

@@

```{r poptimePostImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPostI} Population-time plot before imputation. Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
popTime_post <- popTime(as.data.frame(workingCompleteData),
                        time = "d.time", event = "death")
plot(popTime_post)
```

@@

```{r CBabsolute, echo = TRUE, eval = eval_cs3}
casebaseAbsolute <- absoluteRisk(cb.Model,
                                 time = seq(1, max(completeData$d.time), 0.5),
                                 newdata = t(newData),
                                 s = "lambda.1se", method = "numerical")
```

>>>>

---
documentclass: jss
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 7
fig_height: 6
fig_caption: true
---

@@

library(tidyverse)
library(magrittr)
library(survival)
library(casebase)
library(splines)
library(tibble)
library(glmnet)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- FALSE
eval_cs2 <- FALSE
eval_cs3 <- FALSE
eval_cs4 <- FALSE
```

@@

At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [-@survTaskView]. For the purposes of this article, we restricted our description to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and cumulative incidence curve estimation. By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}. In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with a description of some of the functionalities they offer.

Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche]. Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: \pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package]. On the other hand, \pkg{casebase} provides penalized estimation in the context of parametric hazards. To our knowledge, this is the only package to offer this functionality. \textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}

@@

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks. The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression. \textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}. Could you incorporate some of their discussion, with proper citations? The biggest omission right now is \pkg{cmprsk}, which we actually use below.}

@@

```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group",])
CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group",])
CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group",])

@@

We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [-@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital. Imputation was conducted using the default methods in the \pkg{mice} package in \proglang{R}, available on CRAN. Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest. The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual. A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}. The data was downloaded from the main author's website, and it is also available as part of the \pkg{casebase} package.

@@

Hospital death as a covariate was removed, as this is directly informative of death. The glmnet package(REF) on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase. To visualize the incidence density, we can refer to the post-imputation population-time plot in Figure \ref{fig:ptPostI}.

```{r supportSetUp, echo = FALSE, eval = eval_cs3}
completeData=casebase::support
completeData<-completeData[,-c(3)]
#keep one individual out, at random, for which we will develop an 
#absolute risk curve.
sam <- sample(1:nrow(completeData), 1)
#Set workingData to the remaining individuals in the data
#workingData=completeData[-c(sam),]
workingCompleteData <- as.matrix(sparse.model.matrix(death~ . - d.time, data = completeData))[,-c(1)]
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData <- workingCompleteData[sam,]
x <- workingCompleteData[-c(sam),]
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y <- data.matrix(completeData[-c(sam), c(34, 4)])
```

@@

```{r poptimePostImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPostI} Population-time plot before imputation. Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
popTime_post <- popTime(as.data.frame(completeData),
                        time = "d.time", event = "death")
plot(popTime_post)
```

@@

```{r CBabsolute, echo = TRUE, eval = eval_cs3}

casebaseAbsolute <- absoluteRisk(cb.Model,
                                 time = seq(1, max(completeData$d.time), 0.5),
                                 newdata = t(newData),
                                 s = "lambda.1se", method = "numerical")
```
